---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# rxref

<!-- badges: start -->
[![R-CMD-check](https://github.com/ssmithm/rxref/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/ssmithm/rxref/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

**rxref** is an R package for working with medication identifiers from the
National Library of Medicine's RxNorm and RxClass APIs. It provides tidy,
vectorized helpers for resolving medication names to RxCUIs, expanding
ingredients to product concepts, mapping between RxCUIs and NDCs, retrieving
clinical product attributes, and querying selected drug class relationships.

The package is designed for pharmacoepidemiologic and health services research
workflows where medication exposure definitions often need to move between
ingredient names, RxNorm concepts, National Drug Codes (NDCs), routes, dose
forms, and therapeutic classes.

**rxref** was originally built to support pharmacoepidemiologic research in the
[UF CVmedLab](https://cvmedlab.org/), where we often need reproducible,
transparent medication definitions for prescriptions, pharmacy dispensings,
and claims-based drug exposure measures.

## What rxref can do

Common workflows include:

- resolve free-text medication names to RxNorm ingredient concepts with
  `find_ingredients()` and `resolve()`;
- expand ingredient RxCUIs to product RxCUIs with `products_for_ingredients()`;
- include active-only or active plus historical RxNorm concepts using
  `concept_status`;
- map between RxCUIs and NDCs with `map_to()`, `map_rxcui_to_ndc()`, and
  `map_ndc_to_rxcui()`;
- retrieve product metadata with `get_properties()` and
  `get_clinical_attributes()`;
- filter products by route with `filter_products_by_route()`;
- query selected RxClass relationships with helpers such as `get_atc()`,
  `get_epc()`, and `get_medrt()`;
- use predefined RxNorm term type sets with `product_ttys()` and
  `tty_catalogue()`.

## Installation

You can install the development version of `rxref` from 
[GitHub](https://github.com/) with:

``` r
# install.packages("pak")
# pak::pak("ssmithm/rxref")
```

## Basic Workflow: ingredients to products

This is a basic example for querying product RxCUIs associated with the ingredient lisinopril (a blood pressure-lowering drug):

```{r ex1}
library(rxref)
library(dplyr)

lisinopril_ing <- find_ingredients("lisinopril")

lisinopril_products <- products_for_ingredients(
  lisinopril_ing$rxcui,
  ttys = product_ttys("default"),
  include_combos = TRUE
)

# Resulting table 
lisinopril_products |> 
  dplyr::mutate(name = paste0("`", name, "`")) |> 
  knitr::kable()


# Count of TTYs in resulting table
lisinopril_products |> 
  count(tty, n_ingredients, sort=TRUE)
```

## Active and historical RxNorm concepts

By default, `products_for_ingredients()` searches active RxNorm concepts:

```{r active_historical}
products_for_ingredients(
  "29046",
  ttys = product_ttys("extended_product"),
  concept_status = "active"
) |> knitr::kable()
```

For older prescribing (or dispensing) data, users may need products that are no
longer active in the current RxNorm release. Set `concept_status = "active_and_historical"` 
to include active concepts plus selected historical statuses:

```{r active_historical2}
products_for_ingredients(
  "1369",
  ttys = "SCD",
  include_combos = FALSE,
  concept_status = "active_and_historical"
) |> knitr::kable()
```

Historical searches return additional status and date fields, including
concept_status, active_start_date, active_end_date,
release_start_date, and release_end_date.

**Note**: Historical searches are slower than active-only searches because they require
broader RxNorm status lookups. `rxref` caches repeated API calls where possible,
but active-only queries should be preferred when historical concepts are not
needed.

## Clinical product attributes and route filtering

`get_clinical_attributes()` retrieves product-level attributes such as
ingredients, dose form, dose form group, route, strength, and ingredient count.

```{r clinical_attrs}
attrs <- get_clinical_attributes(lisinopril_products$product_rxcui)

attrs |>
  select(rxcui, related_rxcui, name, strength, dose_form, route, tty)
```

Products can also be filtered by route:

```{r route_filtering}
oral_lisinopril <- filter_products_by_route(
  lisinopril_products,
  route = "ORAL"
)

oral_lisinopril 
```

## Mapping between RxCUIs and NDCs

RxCUIs can be mapped to NDCs (and vice-versa):

```{r rxcui_to_ndc}
ndcs <- map_rxcui_to_ndc("314077")

ndcs
```

The direction-agnostic `map_to()` function can be used when you want to map
between identifier types with one interface:

```{r agnostic_map_to}
map_to("314077", to = "ndc") 
```

## Drug class relationships

`rxref` includes selected helpers for querying RxClass relationships, including
ATC, EPC, and MED-RT classes.

```{r class_rels}
get_atc("29046")
get_epc("29046")
get_medrt("29046")
```

## API use and caching

`rxref` queries public RxNorm and RxClass APIs. The package includes configurable
rate-delay and caching behavior to reduce repeated API requests and to support
longer medication-definition workflows. Note that RxNorm indicates a rate of 
[no more than 20 API requests per second per IP](https://lhncbc.nlm.nih.gov/RxNav/TermsofService.html).

```{r conf}
rxref_conf()
rxref_conf(rate_delay = 0.1)
```

## Notes

RxNorm concepts and NDC mappings can change over time. For reproducible
research, consider recording the date of API access and retaining intermediate
drug-definition outputs used in analyses.

`rxref` is intended to support research workflows and medication-definition
development. Users should review outputs carefully before using them in any 
high-stakes regulatory, clinical, or production settings.
