---
title: "Case Study: GLP-1 and related incretin therapies"
author: "Steven Smith"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteEncoding{UTF-8}
  %\VignetteIndexEntry{Case Study: GLP-1 and related incretin therapies}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

library(rxref)
library(dplyr)

read_rxref_example <- function(file) {
  path <- system.file("extdata", file, package = "rxref")

  if (!nzchar(path)) {
    stop(
      "The example data file '", file, "' was not found. ",
      "You may need to reinstall rxref to restore the bundled vignette data."
    )
  }

  readRDS(path)
}
```

## Case

Suppose we need to identify users of GLP-1 receptor agonists and related
incretin-based therapies from EHR prescribing data, pharmacy claims data,
or both. To accomplish this, we need a reproducible medication definition
that identifies relevant RxNorm product concepts and, when needed,
corresponding National Drug Codes (NDCs).

This case study builds that definition in four steps:

1. define the ingredients of interest;
2. resolve the ingredient names to RxNorm ingredient RxCUIs;
3. expand those ingredients to product-level RxNorm concepts; and
4. map the final product concepts to NDCs.

We first work through these steps explicitly. At the end, we show how
`search_drug()` can provide a more compact version of the same common
workflow.

The displayed results use precomputed data bundled with `rxref`, so
the vignette can be built without requiring access to the live RxNorm API.
The code shown below is the code users can run interactively to generate
the results from the current API.

## Define the ingredient list

For this example, we start with a list of GLP-1 receptor
agonists and related incretin-based therapies:

- exenatide
- liraglutide
- lixisenatide
- dulaglutide
- albiglutide
- semaglutide
- tirzepatide

Tirzepatide is included because many applied studies group it with
GLP-1-based incretin therapies, even though it is technically a dual GIP/GLP-1 
receptor agonist rather than a standalone GLP-1 receptor agonist.

```{r define-ingredients}
glp1_names <- c(
  "semaglutide",
  "exenatide",
  "liraglutide",
  "lixisenatide",
  "dulaglutide",
  "albiglutide",
  "tirzepatide"
)

glp1_names
```

## Step 1: Identify ingredient RxCUIs

`find_ingredients()` resolves free-text drug names to ingredient-level
RxNorm concepts. For this medication definition, we retain TTY = `"IN"`,
which represents the base RxNorm ingredient concept.

```{r identify-ingredients, eval = FALSE}
glp1_ingredients <- find_ingredients(glp1_names) |>
  filter(tty == "IN") |>
  distinct(
    input,
    ingredient_rxcui = rxcui,
    ingredient_name = name,
    ingredient_tty = tty
  )

glp1_ingredients
```

```{r identify-ingredients-data, include = FALSE}
glp1_ingredients <- read_rxref_example("glp1_ings.rds")
```

```{r identify-ingredients-output, echo = FALSE}
glp1_ingredients
```

This step is useful for checking that each study-defined ingredient maps
to the intended RxNorm concept before expanding the definition to
products. For unfamiliar term types, `tty_catalogue()` provides a
description of common RxNorm TTYs.

## Step 2: Expand ingredients to product RxCUIs

Next, `products_for_ingredients()` identifies product concepts that
contain the selected ingredients.

By default, `product_ttys("default")` includes the product-focused TTYs
SCD, SBD, GPCK, and BPCK. These are generally the most useful concepts
when the goal is to identify prescribable or dispensable products and
eventually map them to NDCs.

```{r show-product-ttys}
product_ttys("default")
```

For this example, we retain fixed-dose combination products and use only
currently active RxNorm concepts.

```{r expand-products, eval = FALSE}
glp1_products <- products_for_ingredients(
  glp1_ingredients$ingredient_rxcui,
  ttys = product_ttys("default"),
  include_combos = TRUE,
  concept_status = "active"
)

glp1_products
```

```{r expand-products-data, include = FALSE}
glp1_products <- read_rxref_example("glp1_prods.rds")
```

```{r expand-products-output, echo = FALSE}
glp1_products |>
  arrange(ingredient_rxcui, tty, name) |>
  head(30)
```

### Inspect the product set

Before mapping products to NDCs, it is useful to inspect what was
returned. For example, we can summarize the product TTYs:

```{r inspect-product-ttys}
glp1_products |>
  count(tty, sort = TRUE)
```

Because `include_combos = TRUE`, the product set may also contain
fixed-dose combination products. `n_ingredients` can be used to identify
them:

```{r inspect-combinations}
glp1_products |>
  filter(n_ingredients > 1) |>
  select(ingredient_rxcui, product_rxcui, name, tty, n_ingredients) |>
  arrange(name) |>
  head(20)
```

Whether these products belong in the final exposure definition depends on the 
study. The important point is to make that decision explicit vs. assuming that 
every product containing an ingredient of interest should automatically be retained.

## Step 3: Map product RxCUIs to NDCs

If the exposure data contain NDCs, the next step is to map the selected
product RxCUIs to corresponding NDCs.

Here we request currently active NDC associations and retain NDCs whose
reported status is `"ACTIVE"`.

```{r map-ndcs, eval = FALSE}
glp1_ndc_map <- map_rxcui_to_ndc(
  unique(glp1_products$product_rxcui),
  history = "active",
  status = "ACTIVE"
)
```

```{r map-ndcs-data, include = FALSE}
glp1_ndc_map <- read_rxref_example("glp1_ndc_map.rds")
```

We can then join the NDC mapping back to the product and ingredient
information to create a study-friendly medication list.

```{r build-ndc-list}
glp1_ndcs <- glp1_ndc_map |>
  left_join(
    glp1_products,
    by = c("rxcui" = "product_rxcui")
  ) |>
  left_join(
    glp1_ingredients |>
      select(ingredient_rxcui, ingredient_name),
    by = "ingredient_rxcui"
  ) |>
  distinct(
    ingredient_rxcui,
    ingredient_name,
    product_rxcui = rxcui,
    ndc11,
    ndc_status,
    product_name = name,
    tty
  ) |>
  arrange(ingredient_name, product_rxcui, ndc11)

glp1_ndcs |>
  head(30)
```

At this point, we have two useful study resources:

- `glp1_products`, for data represented with RxCUIs (EHR prescribing data, often); and
- `glp1_ndcs`, for data represented with NDCs (pharmacy fill data, often).

Keeping both tables can be useful when the same medication definition is
applied across multiple data sources.

## Step 4: Perform a few quality checks

Medication-list construction should generally include some inspection of
the final product and NDC sets. Even if you're confident you've done everything
right, RxNorm is occasionally imperfect and sometimes makes weird connections
between ingredients and products, etc...

For example, some basic QC might include a count the number of unique products 
and NDCs represented by each ingredient, to make sure these seem plausible:

```{r qc-counts}
glp1_ndcs |>
  group_by(ingredient_name) |>
  summarise(
    n_products = n_distinct(product_rxcui),
    n_ndcs = n_distinct(ndc11, na.rm = TRUE),
    .groups = "drop"
  ) |>
  arrange(ingredient_name)
```

We can also identify product concepts for which no active NDC was found:

```{r qc-unmapped}
glp1_products |>
  anti_join(
    glp1_ndcs |>
      filter(!is.na(ndc11)) |>
      distinct(product_rxcui),
    by = "product_rxcui"
  ) |>
  select(ingredient_rxcui, product_rxcui, name, tty, n_ingredients) |>
  arrange(name)
```

A product without an NDC is not necessarily an error. RxNorm includes
concepts that do not have a direct current NDC mapping (sometimes no company is
marketing a drug, etc...). The appropriate response depends on the identifiers 
available in the study data and the intended exposure definition.

## A compact alternative with `search_drug()`

Once the underlying workflow is understood, `search_drug()` can provide a
convenient shortcut. It combines ingredient resolution, product
expansion, and optional NDC mapping.

For the same current-concept/current-NDC definition used above:

```{r shortcut-search, eval = FALSE}
glp1_ndcs_shortcut <- search_drug(
  term = glp1_names,
  return = "ndc",
  ndc_status = "ACTIVE",
  concept_status = "active"
)

glp1_ndcs_shortcut
```

```{r shortcut-search-data, include = FALSE}
glp1_ndcs_shortcut <- read_rxref_example("glp1_ndcs_search.rds") |>
  filter(ndc_status == "ACTIVE")
```

```{r shortcut-search-output, echo = FALSE}
glp1_ndcs_shortcut |>
  arrange(product_rxcui, ndc11) |>
  head(30)
```

The shortcut is useful when the desired medication definition matches the
standard workflow. The explicit approach is preferable when users need
to inspect intermediate results, modify the product set, apply
study-specific exclusions, or document exactly how the final medication
definition was developed.

## Adapting the workflow

The choices above define one particular medication list, i.e., GLP1-RAs. Other 
studies may need a different product universe and different needs.

### Exclude fixed-dose combination products

If the study should include only single-ingredient products, set
`include_combos = FALSE` during product expansion:

```{r no-combinations, eval = FALSE}
glp1_products_single <- products_for_ingredients(
  glp1_ingredients$ingredient_rxcui,
  ttys = product_ttys("default"),
  include_combos = FALSE,
  concept_status = "active"
)
```

Alternatively, you might retain combination products initially, inspect
them, and apply more specific study-defined exclusions before mapping to
NDCs.

### Use a broader set of product-related TTYs

The default TTY set is intentionally focused on product concepts that
commonly map to NDCs. A broader product-related set is available when the
research question requires it:

```{r broader-ttys, eval = FALSE}
glp1_products_extended <- products_for_ingredients(
  glp1_ingredients$ingredient_rxcui,
  ttys = product_ttys("extended_product"),
  include_combos = TRUE,
  concept_status = "active"
)
```

Broader TTY sets can be useful for some RxCUI-based searches, but note that not 
all of the additional concept types will map directly to NDCs.

### Include historical RxNorm concepts and NDC associations

Historical RxNorm concepts and historical NDC associations are related
but distinct choices.

To include historical RxNorm product concepts, use
`concept_status = "active_and_historical"`:

```{r historical-products, eval = FALSE}
glp1_products_historical <- products_for_ingredients(
  glp1_ingredients$ingredient_rxcui,
  ttys = product_ttys("default"),
  include_combos = TRUE,
  concept_status = "active_and_historical"
)
```

To retrieve historical NDC associations for those products, use the
`history` argument in `map_rxcui_to_ndc()`:

```{r historical-ndcs, eval = FALSE}
glp1_ndcs_historical <- map_rxcui_to_ndc(
  unique(glp1_products_historical$product_rxcui),
  history = "all"
)
```

The distinction is important:

- `concept_status` determines whether the product search includes
  historical RxNorm concepts.
- `history` determines whether RxCUI-to-NDC mapping uses current or
  historical NDC associations.
- `status` filters NDCs after those associations have been retrieved.

For example, `history = "all", status = "OBSOLETE"` requests historical
NDC associations and then retains NDCs currently reported as obsolete.

`search_drug()` is convenient for current medication-list construction,
but it does not currently expose the NDC `history` argument. For studies
that require historical NDC associations, the explicit
`products_for_ingredients()` plus `map_rxcui_to_ndc()` workflow provides
the necessary control.

## Finalizing the medication definition

Before using a medication list in an analysis, users should consider:

- whether the ingredient list reflects the intended clinical definition;
- whether fixed-dose combination products should be included;
- whether the default product TTYs are sufficient;
- whether route, dose form, or strength restrictions are needed;
- whether the study period requires historical RxNorm concepts or
  historical NDC associations; and
- whether the final product and NDC lists should undergo clinical review.

For reproducible research, save the final medication tables used in the
analysis rather than rebuilding them each time the analysis is run. It is
also useful to record the `rxref` version and the date on which the
medication definition was generated.
