---
title: "Case Study: GLP1 receptor agonists"
author: "Steven Smith"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteEncoding{UTF-8}
  %\VignetteIndexEntry{Case Study: GLP1 receptor agonists}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{r, include = FALSE}
# Vignettes use precomputed example data by default.
# To rebuild examples with live RxNorm/RxClass API calls, set:
# Sys.setenv(RXREF_BUILD_VIGNETTES_ONLINE = "true")

online_env <- identical(
  tolower(Sys.getenv("RXREF_BUILD_VIGNETTES_ONLINE")),
  "true"
)

has_net <- tryCatch({
  requireNamespace("curl", quietly = TRUE) && curl::has_internet()
}, error = function(e) FALSE)

run_live <- online_env && has_net

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

library(rxref)
library(dplyr)

read_rxref_example <- function(file) {
  path <- system.file("extdata", file, package = "rxref")
  if (!nzchar(path)) {
    stop(
      "The example data file '", file, "' was not found. ",
      "Reinstall rxref or rebuild the vignette with ",
      "RXREF_BUILD_VIGNETTES_ONLINE=true."
    )
  }
  readRDS(path)
}
```

## Case

Suppose we have a need to identify all glucagon-like peptide-1 receptor
agonist (GLP1-RA) users from either EHR prescribing data, or pharmacy
claims data (or both). To accomplish this, we need a complete list of
RxCUIs and their corresponding NDCs (when available). We'll go through two 
options to get this list. 

## Option 1: Starting with a complete list of drugs

For this example, we start with a prespecified list of GLP1-RA agents:

-   exenatide

-   liraglutide

-   lixisenatide

-   dulaglutide

-   albiglutide

-   semaglutide

-   tirzepatide


### Identifying Ingredient RxCUIs

Let's turn this into a character vector and use `find_ingredients()` to
gather RxCUIs for TTY = 'IN' (note: you can get a complete list of TTYs
and their description from `tty_catalogue()`).

```{r ings}
glp1.names <- c("semaglutide", "exenatide", "liraglutide", "lixisenatide", "dulaglutide", "albiglutide", "tirzepatide")

if (run_live) {
  glp1.ings <- find_ingredients(glp1.names)
} else {
  glp1.ings <- read_rxref_example("glp1_ings.rds")
}

glp1.ings
```

### Getting Product RxCUIs from Ingredient RxCUIs

Now, let's look for all of the related RxCUIs using
`products_for_ingredients()`. Here, we're going to just focus on the
TTYs that are likely to have NDCs (SCD, SBD, GPCK, BPCK) by using the
function's default TTY.

The default product TTYs are available with `default_product_ttys()`.
These include semantic clinical drugs, semantic branded drugs, generic
packs, and branded packs. A broader set of product-related TTYs is
available with `extended_product_ttys()` or `product_ttys("extended")`.
Users can also supply their own character vector of TTYs.

```{r tty-sets}
default_product_ttys()
extended_product_ttys()
extended_ttys()
```

But, you could supply your own set of TTYs to include instead of using
these two default lists.

Lastly, we're going to include combination products (i.e., if they have
a GLP1-RA plus one or more other drugs), again using the function's
default option of include_combos = TRUE.

```{r prods}
if (run_live) {
  glp1.prods <- products_for_ingredients(glp1.ings$rxcui)
} else {
  glp1.prods <- read_rxref_example("glp1_prods.rds")
}

glp1.prods |> head(30)
```

### Adding in NDCs from Product RxCUIs

Now, let's identify the related NDCs for each RxCUI. Note that not all
RxCUIs will link to NDCs, so some NAs will likely show up in the
resulting file.

```{r ndcs}
if (run_live) {
  glp1.ndc.map <- map_rxcui_to_ndc(unique(glp1.prods$product_rxcui))
} else {
  glp1.ndc.map <- read_rxref_example("glp1_ndc_map.rds")
}

# Join back names/TTYs and de-dupe
glp1_ndcs <- glp1.ndc.map |> 
  dplyr::left_join(glp1.prods, by = c("rxcui" = "product_rxcui")) |> 
  dplyr::distinct(rxcui, ingredient_rxcui, ndc11, name, tty) |> 
  dplyr::arrange(ingredient_rxcui, rxcui, ndc11, name)

glp1.ndcs <- dplyr::left_join(
  x = glp1_ndcs,
  y = glp1.ings |> dplyr::select(rxcui, name) |> dplyr::rename(ingredient = name, ingredient_rxcui = rxcui),
  by = "ingredient_rxcui"
)

glp1.ndcs |> head(30)
```

Voilà, we now have a complete list of GLP1-RAs that can be used to query
against the prescribing data or pharmacy claims data.

## Option 2: starting with the same list of drugs, using search_drug()

This same goal can be accomplished all in one step, if you prefer, using
the `search_drug()` function. Suppose we want to just get NDCs, and we
want to make sure we're capturing *active*, as well as *obsolete* or
*unspecified* NDCs. That is, we don't actually care about the RxCUIs
that don't have a corresponding NDC.

```{r search}

if (run_live) {
  alt.glp1.ndcs <- search_drug(
    term = glp1.names,
    return = "ndc",
    ndc_status = c("ACTIVE", "OBSOLETE", "UNSPECIFIED")
  )
} else {
  alt.glp1.ndcs <- read_rxref_example("alt_glp1_ndc.rds")
}

alt.glp1.ndcs1 <- 
  dplyr::left_join(alt.glp1.ndcs, glp1.prods |> select(-ingredient_rxcui), by = c("product_rxcui" = "product_rxcui")) |> 
  dplyr::distinct(product_rxcui, ingredient_rxcui, ndc11, name, tty) |> 
  dplyr::arrange(ingredient_rxcui, product_rxcui, ndc11, name)
alt.glp1.ndcs2 <- dplyr::left_join(
  x = alt.glp1.ndcs1,
  y = glp1.ings |> dplyr::select(rxcui, name) |> dplyr::rename(ingredient = name, ingredient_rxcui = rxcui),
  by = "ingredient_rxcui")
alt.glp1.ndcs2 
```

Compare that against the original table, filtered for non-NA `ndc11`
values:

```{r check}
glp1.ndcs |> filter(!is.na(ndc11))
```
