---
title: "Case Study: Building route-specific RxCUI and NDC lists"
author: "Steven Smith"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Case Study: Building route-specific RxCUI and NDC lists}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

library(rxref)
library(dplyr)

read_rxref_example <- function(file) {
  path <- system.file("extdata", file, package = "rxref")

  if (!nzchar(path)) {
    stop(
      "The example data file '", file, "' was not found. ",
      "You may need to reinstall rxref to restore the bundled vignette data."
    )
  }

  readRDS(path)
}
```

## Case

A common use case for `rxref` is to develop a list of RxNorm product concepts 
(RxCUIs) and National Drug Codes (NDCs) for a medication class. However, 
sometimes the ingredient alone is not sufficient to define the exposure.

Suppose we want to identify **oral beta-blocker products** for an
outpatient pharmacoepidemiologic study. Several beta-blockers are
available via multiple routes of administration. For example:

- metoprolol, atenolol, labetalol, propranolol, and sotalol have oral and
  injectable products;
- esmolol is primarily used as an injectable product; and
- timolol, betaxolol, and related drugs may have ophthalmic products.

If we simply expand beta-blocker ingredients to all products and map them
to NDCs, we would capture products outside the intended exposure
definition.

This case study builds an oral beta-blocker medication list in five
steps:

1. define the ingredients of interest;
2. resolve ingredient names to RxNorm ingredient RxCUIs;
3. expand the ingredients to product-level RxNorm concepts;
4. inspect and filter products by route; and
5. map the final oral products to active NDCs.

The displayed results use precomputed data bundled with `rxref`, so
the vignette can be built without requiring access to the live RxNorm API.
The code shown below is the code users can run interactively to generate
the results from the current API.

## Define the ingredient list

We begin with a curated list of beta-blocker ingredients.

```{r define}
beta_blocker_names <- c(
  "acebutolol",
  "atenolol",
  "betaxolol",
  "bisoprolol",
  "carvedilol",
  "labetalol",
  "metoprolol",
  "nadolol",
  "nebivolol",
  "penbutolol",
  "pindolol",
  "propranolol",
  "sotalol",
  "timolol"
)

beta_blocker_names
```

A curated ingredient list is often appropriate when the study definition
is known in advance. `rxref` also provides RxClass helpers such as
`find_classes()` and `get_class_members()` when users want to begin from
a source vocabulary class rather than a prespecified list.

## Step 1: Identify ingredient RxCUIs

Use `find_ingredients()` to resolve each name to an RxNorm ingredient
concept. Here we retain TTY = `"IN"`, representing the base ingredient
concept.

```{r resolve, eval = FALSE}
bb_ingredients <- find_ingredients(beta_blocker_names) |>
  filter(tty == "IN") |>
  distinct(
    input,
    ingredient_rxcui = rxcui,
    ingredient_name = name,
    ingredient_tty = tty
  )

bb_ingredients
```

```{r resolve-data, include = FALSE}
bb_ingredients <- read_rxref_example("bb_ingredients.rds")
```

```{r resolve-output, echo = FALSE}
bb_ingredients
```

This is a useful checkpoint: before expanding to hundreds of product
concepts, verify that each study-defined name resolved to the intended
ingredient.

## Step 2: Expand ingredients to product RxCUIs

Next, use `products_for_ingredients()` to identify product-level RxNorm
concepts containing the selected ingredients.

For this example, we use the default product-focused TTYs, retain
fixed-dose combination products, and initially keep products from all
routes.

```{r expand, eval = FALSE}
bb_products <- products_for_ingredients(
  bb_ingredients$ingredient_rxcui,
  ttys = product_ttys("default"),
  include_combos = TRUE,
  concept_status = "active"
) |>
  # rejoin the ingredients list so you know what maps to what
  left_join(bb_ingredients, by = "ingredient_rxcui")

bb_products
```

```{r expand-data, include = FALSE}
bb_products <- read_rxref_example("bb_products.rds")
```

```{r expand-output, echo = FALSE}
bb_products |>
  arrange(ingredient_name, tty, name) |>
  head(30)
```

At this stage, the product table intentionally contains products from all
routes. Keeping route filtering separate for the moment makes it easier
to inspect what would otherwise be included in the medication
definition.

Because `include_combos = TRUE`, fixed-dose combination products may
also be present. We'll return to that later.

## Step 3: Inspect route and dose-form information

`get_clinical_attributes()` derives clinically useful information from
the product concepts, including route, dose form, dose-form group,
ingredient count, and brand/generic indicators.

```{r inspect-routes, eval = FALSE}
bb_attributes <- get_clinical_attributes(
  unique(bb_products$product_rxcui)
) |>
  rename(product_rxcui = rxcui)

bb_attributes
```

```{r inspect-routes-data, include = FALSE}
bb_attributes <- read_rxref_example("bb_attrs.rds")
```

A quick summary shows why route filtering is needed:

```{r inspect-routes-output}
bb_attributes |>
  count(route, dose_form_group, sort = TRUE)
```

The exact mix of routes depends on the current RxNorm concepts returned
for the selected ingredients, but the important point is that an
ingredient-based search can identify products that do not correspond to
the intended route of exposure.

It can also be useful to inspect individual products associated with
non-oral routes:

```{r inspect-nonoral}
bb_products |>
  left_join(
    bb_attributes |>
      select(product_rxcui, route, dose_form, dose_form_group),
    by = "product_rxcui"
  ) |>
  filter(is.na(route) | route != "ORAL") |>
  select(
    ingredient_name,
    product_rxcui,
    name,
    tty,
    route,
    dose_form,
    dose_form_group
  ) |>
  arrange(ingredient_name, route, name) |>
  head(30)
```

This inspection step is especially useful when working with a medication
class for which route heterogeneity is clinically important.

## Step 4: Filter to oral products

Once we have confirmed that the unfiltered product set contains multiple
routes, use `filter_products_by_route()` to retain only oral products.

```{r filter, eval = FALSE}
bb_oral_products <- bb_products |>
  filter_products_by_route(route = "ORAL")

bb_oral_products
```

```{r filter-data, include = FALSE}
bb_oral_products <- read_rxref_example("bb_oral_products.rds")
```

```{r filter-output, echo = FALSE}
bb_oral_products |>
  arrange(ingredient_name, tty, name) |>
  head(30)
```

The returned table preserves the product-level information and appends
summarized route, dose-form, and dose-form-group fields.

```{r filter-summary}
bb_oral_products |>
  count(route, dose_form_group, sort = TRUE)
```

Route filtering depends on clinical attribute information available for a
product RxCUI. A product without a route match is not retained, so
route-specific medication lists should always be reviewed before they are
used in an analysis.

## Step 5: Map oral product RxCUIs to active NDCs

After restricting the RxCUI definition to oral products, map those
products to NDCs.

For a current medication list, we use current NDC associations and retain
NDCs whose reported status is `"ACTIVE"`.

```{r map, eval = FALSE}
bb_oral_ndc_map <- map_rxcui_to_ndc(
  unique(bb_oral_products$product_rxcui),
  history = "active",
  status = "ACTIVE"
)

bb_oral_ndcs <- bb_oral_ndc_map |>
  left_join(
    bb_oral_products,
    by = c("rxcui" = "product_rxcui")
  ) |>
  rename(product_rxcui = rxcui) |>
  distinct()

bb_oral_ndcs
```

```{r map-data, include = FALSE}
bb_oral_ndcs <- read_rxref_example("bb_oral_ndcs.rds")
```

```{r map-output, echo = FALSE}
bb_oral_ndcs |>
  select(
    ingredient_name,
    product_rxcui,
    name,
    tty,
    route,
    dose_form,
    ndc11,
    ndc_status
  ) |>
  arrange(ingredient_name, product_rxcui, ndc11) |>
  head(30)
```

At this point, we have two useful study resources:

- `bb_oral_products`, containing the route-restricted RxCUI definition;
  and
- `bb_oral_ndcs`, containing the corresponding active NDC mappings.

Keeping both can be useful when the medication definition will be applied
to data sources that use different drug identifiers.

## Quality-control checks

Route-specific medication definitions benefit from several simple quality
checks before they are finalized.

### Confirm the observed routes

```{r qc-routes}
bb_oral_products |>
  count(route, dose_form_group, sort = TRUE)
```

For this definition, every retained product should have an oral route
assignment.

### Look for non-oral terminology in product names

A name-based check is not a substitute for structured route filtering,
but it is a useful secondary screen.

```{r qc-nonoral-names}
bb_oral_products |>
  filter(grepl(
    "Injection|Injectable|Ophthalmic|Topical|Transdermal|Nasal|Inhalation",
    name,
    ignore.case = TRUE
  )) |>
  select(
    ingredient_name,
    product_rxcui,
    name,
    tty,
    route,
    dose_form
  )
```

Unexpected rows should be reviewed rather than automatically removed,
because product naming conventions do not always map perfectly to route. And,
again, sometimes RxNorm has mistakes in the source data that will eventually
get corrected.

### Review combination products

Because this example retained combination products, inspect them
explicitly:

```{r qc-combinations}
bb_oral_products |>
  filter(n_ingredients > 1) |>
  select(
    ingredient_name,
    product_rxcui,
    name,
    tty,
    n_ingredients,
    route,
    dose_form
  ) |>
  arrange(ingredient_name, name)
```

For example, beta-blocker/thiazide combination products may be appropriate
for a broad antihypertensive exposure definition but inappropriate for a
study specifically evaluating beta-blocker monotherapy.

### Check products without active NDCs

Some valid RxNorm product concepts may not map to a current active NDC.

```{r qc-without-ndcs}
bb_oral_products |>
  anti_join(
    bb_oral_ndcs |>
      filter(!is.na(ndc11)) |>
      distinct(product_rxcui),
    by = "product_rxcui"
  ) |>
  select(
    ingredient_name,
    product_rxcui,
    name,
    tty,
    route,
    dose_form
  ) |>
  arrange(ingredient_name, name)
```

Whether products without active NDCs matter depends on the identifiers
available in the study data.

### Summarize the final definition

Finally, summarize the number of unique products and NDCs associated with
each ingredient:

```{r qc-counts}
bb_oral_ndcs |>
  group_by(ingredient_name) |>
  summarise(
    n_products = n_distinct(product_rxcui),
    n_ndcs = n_distinct(ndc11, na.rm = TRUE),
    .groups = "drop"
  ) |>
  arrange(ingredient_name)
```

## A compact alternative with `search_drug()`

Once the explicit workflow is understood, `search_drug()` can combine
ingredient resolution, product expansion, route filtering, and NDC
mapping in a single call.

Using `return = "both"` is particularly useful here because the product
table and the NDC table have different grains.

```{r shortcut, eval = FALSE}
bb_oral <- search_drug(
  beta_blocker_names,
  return = "both",
  route = "ORAL",
  ndc_status = "ACTIVE",
  include_combos = TRUE,
  concept_status = "active"
)

bb_oral$products
bb_oral$ndcs
```

```{r shortcut-data, include = FALSE}
bb_oral <- read_rxref_example("bb_oral_both_search.rds")
```

The result is a list with separate product and NDC tables:

```{r shortcut-names}
names(bb_oral)
```

```{r shortcut-products}
bb_oral$products |>
  arrange(ingredient_rxcui, tty, name) |>
  head(20)
```

```{r shortcut-ndcs}
bb_oral$ndcs |>
  arrange(ingredient_rxcui, product_rxcui, ndc11) |>
  head(20)
```

The shortcut is useful when the standard workflow matches the study
definition. The explicit approach remains preferable when users need to
inspect route heterogeneity before filtering, modify the product set, or
document study-specific inclusion and exclusion decisions.

## Adapting the workflow

The example above represents one particular route-specific medication
definition. Several common modifications are possible.

### Filter by route during product expansion

When users already know the intended route and do not need to inspect the
unfiltered product universe first, `products_for_ingredients()` can apply
the route restriction directly:

```{r direct-route-filter, eval = FALSE}
bb_oral_products_direct <- products_for_ingredients(
  bb_ingredients$ingredient_rxcui,
  ttys = product_ttys("default"),
  route = "ORAL",
  include_combos = TRUE,
  concept_status = "active"
)
```

This uses the same route-filtering infrastructure but provides a more
compact workflow.

### Exclude fixed-dose combination products

If the study should include beta-blocker single drug entities (i.e., not 
combination products) only, set `include_combos = FALSE`:

```{r exclude-combinations, eval = FALSE}
bb_oral_single_products <- products_for_ingredients(
  bb_ingredients$ingredient_rxcui,
  ttys = product_ttys("default"),
  route = "ORAL",
  include_combos = FALSE,
  concept_status = "active"
)
```

Alternatively, retain combination products initially, review them, and
apply study-specific exclusions after product expansion.

### Include historical RxNorm concepts and NDC associations

For older study periods, users may need concepts that are no longer
active in the current RxNorm release.

Historical RxNorm product concepts can be included with:

```{r historical-products, eval = FALSE}
bb_products_historical <- products_for_ingredients(
  bb_ingredients$ingredient_rxcui,
  ttys = product_ttys("default"),
  include_combos = TRUE,
  concept_status = "active_and_historical"
)

bb_oral_products_historical <- bb_products_historical |>
  filter_products_by_route(route = "ORAL")
```

`filter_products_by_route()` can use historical RxCUI metadata as a
fallback when the product table contains non-active concepts. Even so,
historical route assignments deserve careful review because clinical
attribute information may be less complete for older concepts.

Historical NDC associations are a separate choice. Retrieve them with the
`history` argument in `map_rxcui_to_ndc()`:

```{r historical-ndcs, eval = FALSE}
bb_oral_ndcs_historical <- map_rxcui_to_ndc(
  unique(bb_oral_products_historical$product_rxcui),
  history = "all"
)
```

The distinction is important:

- `concept_status` determines whether product expansion includes
  historical RxNorm concepts;
- `history` determines whether RxCUI-to-NDC mapping uses current or
  historical NDC associations; and
- `status` filters NDCs after those associations have been retrieved.

`search_drug()` is convenient for current route-specific medication lists,
but it does not currently expose the NDC `history` argument. For studies
requiring historical NDC associations, use the explicit
`products_for_ingredients()` plus `map_rxcui_to_ndc()` workflow.

## Finalizing a route-specific medication definition

Before using a route-specific medication list in an analysis, consider:

- whether the ingredient list matches the intended clinical definition;
- whether the selected route is represented correctly across products;
- whether combination products should be included;
- whether the default product TTYs are sufficient;
- whether products without route information require manual review;
- whether products without active NDCs matter for the source data; and
- whether the study period requires historical RxNorm concepts or NDC
  associations.

For reproducible research, save the final product and NDC tables used in
the analysis rather than rebuilding them each time. It is also useful to
record the `rxref` version and the date on which the medication definition
was generated.
